Towards Verifying the Bitcoin-S Library

### 2019-07-06 paper/paper.tex
final touches

\begin{abstract}
  We try to verify some properties of the bitcoin-s library, a Scala
  implementation of parts of the Bitcoin protocol. We use the
  Stainless verifier which supports programs in a fragment of Scala
  called \emph{Pure Scala}. Since bitcoin-s is not written in this
  fragment, we extract the relevant code from it and perform a series
  of equivalent transformations until we arrive at code that we
  successfully verify. In that process we find and fix two bugs in
  bitcoin-s.

@@

\paragraph{The Stainless Verifier.} Stainless is the successor of the
Leon verifier
\cite{DBLP:conf/ecoop/BlancKKS13,DBLP:conf/pldi/VoirolKK15,DBLP:conf/pldi/BlancK15}
and is developed at EPF Lausanne \cite{Stainless:github}. It is
intended to be used by programmers without training in formal
verification. To facilitate that, it accepts specifications written in
the programming language itself (Scala). Also, it focusses on
counterexample finding in addition to proving
correctness. Counterexamples are useful to programmers while
correctness proofs are not -- correctness is obvious or does not hold,
and often both at the same time.

The example in Figure~\ref{fig:factorial} adapted from the Stainless
documentation \cite{Stainless:documentation} shows how the verifier is
used. Notice how a precondition is specified using \emph{require} and
a postcondition using \emph{ensuring}.

\begin{figure}
\begin{lstlisting}[style=scala]
  def factorial(n: Int): Int = {
      require(n >= 0)
      if (n == 0) {
        1
      } else {
        n * factorial(n - 1)
      }
  } ensuring(res => res >= 0)
\end{lstlisting}
	\caption{Factorial function with specification}
	\label{fig:factorial}
\end{figure}
Our function does not satisfy the specification. An overflow in the
32-bit Int leads to a negative result for the input 17, as Stainless
reports in Figure~\ref{fig:failed}. Changing the type Int to BigInt
will result in a successful verification.

@@

\paragraph{The Pure Scala Fragment.} The Scala fragment supported by
Stainless is described in the Stainless documentation
\cite{Stainless:documentation} in the section
\href{https://epfl-lara.github.io/stainless/purescala.html}{Pure
  Scala}.

It comprises algebraic data types in the form of abstract classes,
case classes and case objects, objects for grouping classes and
functions, boolean expressions with short-circuit interpretation,
generics with invariant type parameters, default values of function
parameters, pattern matching, local and anonymous classes and more.
In addition to Pure Scala Stainless also supports some imperative
features, such as using a (mutable) variable in a local scope of a
function and while loops. They turn out not to be relevant for the
current work.

What will turn out to be more relevant for us are the Scala features
which Stainless does not support, such as: arbitrary class
definitions, inheritance by objects, abstract type members, and inner
classes in case objects.

Also, Stainless has its own library of some core data types and
functions which are mapped to corresponding data types and functions
inside of the SMT solver that Stainless ultimately relies on. Those
data types in general do not have all the methods of the Scala data
types. For example, the BigInt type in Scala has methods for bitwise
operations while the BigInt type in Stainless does not.

@@

@@

A crucial function for the verification of the no-inflation property
is the \texttt{checkTransaction} function shown in
Figure~\ref{fig:checktrans}. Given a transaction it returns true if
some basic checks succeed, otherwise false. For example, one of those
checks is that both the list of inputs and list of outputs need to be
non-empty.

\paragraph{Creating a Transaction.} The code in this subsection is
adapted from the bitcoin-s documentation. To create a transaction, we
first need some coins -- an unspent transaction output. We could load
an actual unspent transaction output from the bitcoin network, but we
create one manually in order to see this process. So we first create
an (invalid) transaction with one output in Figure~\ref{fig:prevtx}.

@@

We first create a keypair, then a lock script with its public key,
then the amount of satoshis, then a transaction output (utxo) for that
amount and locked with that script. Finally we create a transaction
with that output and no inputs. Of course, that is not a valid
transaction, because it creates coins out of nothing. In particular,
\texttt{checkTransaction(prevTx)} returns false, simply because the list of
inputs is empty.

@@

Now that we have a transaction output, we create a transaction to
spend it in Figure~\ref{fig:tx}. First, we need a reference to an
output of a previous transaction, here called
\texttt{outPoint}. Second, we add some information on how to spend
that output, in particular, how to sign the transaction. Now we
assemble the list of unspent transaction outputs (utxos), in our case
just one.

We then set the amount of satoshis that we want to spend. The Int64
class aims to emulate a C data type in Scala, and we will look at it
more closely in the next section.

We then create a lock script (\textit{destinationSPK}) to receive the
coins, create our list of transaction outputs (\texttt{destinations}),
define the fee rate and set some bitcoin network parameters.

@@

\paragraph{A Bug in the checkTransaction Function.} Note lines 15-17
of the checkTransaction function in Figure~\ref{fig:checktrans}. Here,
the value prevOutputTxIds gathers a list of all transaction identifiers
referenced by the inputs of the current transaction. If the size of
this list is the same as the size of this list with duplicates
removed, we know that no transaction has been referenced twice. This
prevents a transaction from spending two different outputs of the same
previous transaction. The check is too strict: checkTransaction
returns false for valid transactions.

The fix is simple: we perform the duplicate check on the
TransactionOutPoint instances instead of on their transaction
identifiers. Note that TransactionOutPoint is a case class and thus
its notion of equality is just what we need: equality of of both the
txId and the output index.

@@

\paragraph{An Attempt at Verification.} Naively trying Stainless on
the entire bitcoin-s codebase results in many errors -- as was to be
expected. We tried to extract only the code relevant to the
no-inflation-property and to verify that. However, even the extracted
code has more than 1500 lines and liberally uses Scala features
outside of the supported fragment. We tried to transform the code
into the supported fragment, but quickly realized that a better
approach is to first verify a simpler property with less code involved
and later come back to the no-inflation property with more
experience. So we now turn to the addition-with-zero property.

@@

It is of course a crucial property we are verifying here: if zero
satoshis were credited to your account, you would not want your
balance to change! It is also the simplest meaningful property to
verify that we can think of. However, the code involved in performing
the addition of two satoshi amounts in bitcoin-s is non-trivial. The
reason for that is a peculiarity of consensus code: agreement with the
majority is more important than correctness, whatever correctness
might mean. The most widely used bitcoin implementation by far is the
reference implementation Bitcoin Core, written in C++. For consensus
code, bitcoin-s has little choice but to be in strict agreement with
the reference implementation. To achieve that, it implements C-like
data types in Scala and then implements functionality using those
C-like data types. For example, the Satoshis class, which represents
an amount of satoshis, is implemented using the class Int64 which
aims to represent the C-type \texttt{int64\_t}.

@@

\paragraph{Extracting the Relevant Code} The relevant code for the
addition of satoshis is in two files: CurrencyUnits.scala and
NumberType.scala. From those files we removed the majority of the code
because it is not needed for the verification of our property. For
example, we removed all number types except for Int64 (so Int32,
UInt64, etc.) because they are not used. We also removed the
superclasses Factory and NetworkElement of CurrencyUnit and Number,
respectively, because the inherited members are not used. We further
removed all binary operations on Number that are not used, like
subtraction and multiplication. The extracted code is shown in
Figure~\ref{fig:numbertype} and Figure~\ref{fig:currencyunits}.

\paragraph{A Bug in the checkResult Function.} Note the checkResult
function on line 12 and the value andMask on line 23 of
NumberType.scala. The function is intended to catch overflows by
performing a bitwise conjunction of its argument with andMask and
comparing the result with the argument. However, because of the way
Java BigIntegers are represented \cite{wikipedia:twocomp} and because
bitwise operations implicitly perform a sign extension
\cite{java:bigint} on the shorter operand, the function does not
actually catch overflows.

While this is a potentially serious bug, it turns out that checkResult
is only ever called inside a constructor call for a number type (like
Int64) which contains the intended range check, see lines 32-35. The
checkResult function thus can, and should, be removed entirely. The
bitcoin-s developers have acknowledged the bug and we submitted a pull
request to fix it \cite{BitcoinS:pull565}.

\paragraph{Transforming the Code.} We now turn to the list of Scala
features used by the extracted code which are not supported by
Stainless and how to transform the code into the supported fragment.
All transformations are equivalent in the sense that if the
addition-with-zero property holds for the transformed code, then it
also holds for the code before the transformation.

\paragraph{Inheriting Objects.} In both files we have objects
extending the BaseNumbers trait, on lines 30 and 23 respectively,
which Stainless does not support. We simply turn those objects into
case objects. That transformation is equivalent. Case objects have
various additional properties (for example, being serializable) but
none of our code depends on the absence of those.

\paragraph{Abstract Type Members.} In CurrencyUnits.scala on line 6
there is an abstract type that is not supported. Note that we can not
simply replace it by a (supported) type parameter since the
CurrencyUnit class uses one of its implementing classes:
Satoshis. Since the Satoshis class overrides A with Int64 anyway, we
just remove the abstract type declaration and replace A by Int64
everywhere.

\paragraph{Non-Literal BigInt Constructor Argument.} In
CurrencyUnits.scala on line 18 the BigInt constructor is called with a
non-literal argument. As described before, the types in the Stainless
library are more restricted than their Scala library counterparts. In
particular, the Stainless BigInt constructor is restricted to literal
arguments. So we simply replace \texttt{toLong} by
\texttt{underlying.toBigInt}: instead of converting the underlying
Int64 (which in turn has an underlying BigInt) to Long and then back
to BigInt we simply directly return the BigInt. This is an equivalent
transformation: the only thing that might go wrong in the detour via
\texttt{Long} is that the underlying BigInt does not fit into a
Long. However, the only constructor of Int64Impl ensures exactly that
and all functions producing Int64 do so via this constructor.

\paragraph{Self-Reference in Type Parameter Bound.} In
NumberTypes.scala both on lines 3 and 19 is a class with a type
parameter and a type boundary that contains that type parameter
itself. Stainless does not currently support such self-referential
type boundaries. We opened an issue \cite{Stainless:issue519} on the
Stainless repository and the developers have targeted version 0.4 to
support self-referential type boundaries. Since our code only uses
Number with type parameter T instantiated to Int64, we just remove the
type parameter declaration and replace all its occurrences by Int64.

\paragraph{Missing Member \texttt{bigInteger} in BigInt.} In
NumberType on line 6 there is a reference to \texttt{bigInteger}. The
Scala BigInt class is essentially a wrapper around
java.math.BigInteger. BigInt has a member bigInteger which is the
underlying instance of the Java class. The Java class has a method
longValueExact which returns a long only if the BigInteger fits into a
long, otherwise throws exception. Stainless does not support Java
classes and in particular its BigInt has no member
bigInteger. However, our code does not call toLong anymore, so we just
remove it.

\paragraph{Type Members.} In NumberType.scala there is a type member
on line 4. Our version of Stainless (0.1) does not support type
members. We just remove the declaration and replace all occurrences of
A with BigInt, since A is never overwritten in an implementing class.
Note that in the meantime Stainless has implemented support for type
members \cite{Stainless:pull470}. Since version 0.2 verification
should succeed without this change.

\paragraph{Missing Bitwise-And Method on BigInt.} Contrary to Scala
BigInt, the Stainless BigInt class does not support bitwise
operations, in particular not the \&-method used in NumberType.scala
on line 13. However, as described above, the checkResult function is
both broken and redundant, so we remove it and all calls to it.

@@

\paragraph{Missing Implicit Long to BigInt Conversion.} The Scala
BigInt class has implict conversions from Long which NumberType.scala
uses on lines 32 and 34 and they are missing in the Stainless
BigInt. Since a BigInt constructor with a Long argument is also
missing, we replace the Long literals by an explicit call to the
BigInt constructor with a literal string argument,
e.g. \texttt{BigInt("-9223...5808")}.

@@

@@

\todo{reference to verified code}

\todo{say which version of stainless
  we used, which version of bitcoin-s:
  91633375a37a170ba260ff4b443582aca4aa0c77 }

\todo{code in text: italic or courier?}

>>>>

\begin{abstract}
  We try to verify properties of the bitcoin-s library, a Scala
  implementation of parts of the Bitcoin protocol. We use the
  Stainless verifier which supports programs in a fragment of Scala
  called \emph{Pure Scala}. Since bitcoin-s is not written in this
  fragment, we extract the relevant code from it and perform a series
  of equivalent transformations until we arrive at code that we
  successfully verify. In that process we find and fix two bugs in
  bitcoin-s.

@@

\paragraph{The Stainless Verifier.} Stainless is the successor of the
Leon verifier and is developed at EPF Lausanne
\cite{DBLP:conf/ecoop/BlancKKS13,DBLP:conf/pldi/VoirolKK15,DBLP:conf/pldi/BlancK15}. It
is intended to be used by programmers without training in formal
verification. To facilitate that, it accepts specifications written in
the programming language itself (Scala). Also, it focusses on
counterexample finding in addition to proving
correctness. Counterexamples are useful to programmers while
correctness proofs are not -- correctness is obvious or does not hold,
and often both at the same time.

The example in Figure~\ref{fig:factorial} adapted from the Stainless
documentation \cite{Stainless:documentation} shows how the verifier is
used. Note how a precondition is specified using \texttt{require} and
a postcondition using \texttt{ensuring}. Our function does not satisfy
the specification. An overflow in the 32-bit integer type leads to a
negative result for the input 17, as Stainless reports in
Figure~\ref{fig:failed}. Changing the type \texttt{Int} to
\texttt{BigInt} will result in a successful verification.

\begin{figure}
\begin{lstlisting}[style=scala]
  def factorial(n: Int): Int = {
      require(n >= 0)
      if (n == 0) {
        1
      } else {
        n * factorial(n - 1)
      }
  } ensuring(res => res >= 0)
\end{lstlisting}
	\caption{Factorial function with specification}
	\label{fig:factorial}
\end{figure}

@@

\paragraph{The Pure Scala Fragment.} The Scala fragment supported by
Stainless comprises algebraic data types in the form of abstract
classes, case classes and case objects, objects for grouping classes
and functions, boolean expressions with short-circuit interpretation,
generics with invariant type parameters, pattern matching, local and
anonymous classes and more.  In addition to Pure Scala Stainless also
supports some imperative features, such as using a (mutable) variable
in a local scope of a function and while loops. They turn out not to
be relevant for our current work.

What will turn out to be more relevant for us are the Scala features
which Stainless does not support, such as: inheritance by objects,
abstract type members, and inner classes in case objects. Also,
Stainless has its own library of some core data types and functions
which are mapped to corresponding data types and functions inside of
the SMT solver that Stainless ultimately relies on. Those data types
in general do not have all the methods of the Scala data types. For
example, the \texttt{BigInt} type in Scala has methods for bitwise
operations while the \texttt{BigInt} type in Stainless does not.

@@

The \texttt{checkTransaction} function shown in
Figure~\ref{fig:checktrans} is crucial for the verification of the
no-inflation property. Given a transaction it returns true if some
basic checks succeed, otherwise false. For example, one of those
checks is that both the list of inputs and list of outputs need to be
non-empty.

\paragraph{Creating a Transaction.} To create a transaction, we first
need some coins -- an unspent transaction output. We could load an
actual unspent transaction output from the bitcoin network, but we
create one manually in order to see this process. So we first create
an (invalid) transaction with one output in Figure~\ref{fig:prevtx}.

@@

@@

We first create a keypair, then a lock script with its public key,
then the amount of satoshis, then a transaction output (\texttt{utxo})
for that amount and locked with that script. Finally we create a
transaction with that output and no inputs. Of course, that is not a
valid transaction, because it creates coins out of nothing. In
particular, \texttt{checkTransaction(prevTx)} returns false, simply
because the list of inputs is empty.

@@

Now that we have a transaction output, we create a transaction to
spend it in Figure~\ref{fig:tx}. First, we need a reference to an
output of a previous transaction, here called
\texttt{outPoint}. Second, we add some information on how to spend
that output, in particular, how to sign the transaction. Now we
assemble the list of unspent transaction outputs (\texttt{utxos}), in our case
just one.

We then set the amount of satoshis that we want to spend. The \texttt{Int64}
class aims to emulate a C data type in Scala, and we will look at it
more closely in the next section.

We then create a lock script (\texttt{destinationSPK}) to receive the
coins, create our list of transaction outputs (\texttt{destinations}),
define the fee rate and set some bitcoin network parameters.

@@

\paragraph{A Bug in the checkTransaction Function.} Note lines 15-17
in Figure~\ref{fig:checktrans}. Here, the value
\texttt{prevOutputTxIds} gathers a list of all transaction identifiers
referenced by the inputs of the current transaction. If the size of
this list is the same as the size of this list with duplicates
removed, we know that no transaction has been referenced twice. This
prevents a transaction from spending two different outputs of the same
previous transaction. The check is too strict:
\texttt{checkTransaction} returns false for valid transactions.

The fix is simple: we perform the duplicate check on the
\texttt{TransactionOutPoint} instances instead of on their transaction
identifiers. Note that \texttt{TransactionOutPoint} is a case class
and thus its notion of equality is just what we need: equality of of
both the transaction identifier and the output index.

@@

\paragraph{An Attempt at Verification.} Naively trying Stainless on
the entire bitcoin-s codebase results in many errors -- as was to be
expected. We tried to extract only the code relevant to the
no-inflation-property and to verify that. However, even the extracted
code has more than 1500 lines and liberally uses Scala features
outside of the supported fragment. We tried to transform the code into
the supported fragment, but quickly realized that a better approach is
to first verify a simpler property depending on less code and later
come back to the no-inflation property with more experience. So we now
turn to the addition-with-zero property.

@@

It is of course a crucial property we are verifying here: if zero
satoshis were credited to your account, you would not want your
balance to change! It is also the simplest meaningful property to
verify that we can think of. However, the code involved in performing
the addition of two satoshi amounts in bitcoin-s is non-trivial. The
reason for that is a peculiarity of consensus code: agreement with the
majority is more important than correctness, whatever correctness
might mean. The most widely used bitcoin implementation by far is the
reference implementation Bitcoin Core, written in C++. For consensus
code, bitcoin-s has little choice but to be in strict agreement with
the reference implementation. To achieve that, it implements C-like
data types in Scala and then implements functionality using those
C-like data types. For example, the Satoshis class, which represents
an amount of satoshis, is implemented using the class \texttt{Int64} which
aims to represent the C-type \texttt{int64\_t}.

@@

\paragraph{Extracting the Relevant Code} The relevant code for the
addition of satoshis is in two files: CurrencyUnits.scala and
NumberType.scala. From those files we removed the majority of the code
because it is not needed for the verification of our property. For
example, we removed all number types except for \texttt{Int64} (so
\texttt{Int32}, \texttt{UInt64}, etc.) because they are not used. We
also removed the superclasses \texttt{Factory} and
\texttt{NetworkElement} of \texttt{CurrencyUnit} and \texttt{Number},
respectively, because the inherited members are not used. We further
removed all binary operations on \texttt{Number} that are not used,
like subtraction and multiplication. The extracted code is shown in
Figure~\ref{fig:numbertype} and Figure~\ref{fig:currencyunits}.

\paragraph{A Bug in the checkResult Function.} Note the
\texttt{checkResult} function on line 12 and the value
\texttt{andMask} on line 23 of NumberType.scala. The function is
intended to catch overflows by performing a bitwise conjunction of its
argument with \texttt{andMask} and comparing the result with the
argument. However, because of the way Java BigIntegers are represented
\cite{wikipedia:twocomp} and because bitwise operations implicitly
perform a sign extension \cite{java:bigint} on the shorter operand,
the function does not actually catch overflows.

While this is a potentially serious bug, it turns out that
\texttt{checkResult} is only ever called inside a constructor call for
a number type which contains the intended range check, see lines
32-35. The \texttt{checkResult} function thus can, and should, be
removed entirely. The bitcoin-s developers have acknowledged the bug
and we submitted a pull request to fix it \cite{BitcoinS:pull565}.

% For further development, this raises the question: if the goal of the
% \texttt{Int64} type is to emulate \texttt{int64\_t} then why does it
% prevent overflows? To achieve strict agreement with Bitcoin Core, a
% better approach might be to remove overflow checking from the data
% type and to add it in exactly those places where it happens in Bitcoin
% Core.

\paragraph{Transforming the Code.} We now turn to the list of Scala
features used by the extracted code which are not supported by
Stainless and how to transform the code into the supported fragment.
All transformations are \emph{equivalent} in the sense that if the
addition-with-zero property holds for the transformed code, then it
also holds for the code before the transformation.

\paragraph{Inheriting Objects.} In both files we have objects
extending the BaseNumbers trait, on lines 30 and 23 respectively,
which Stainless does not support. We simply turn those objects into
case objects. That transformation is equivalent: case objects have
various additional properties (for example, being serializable) but
none of our code depends on the absence of those.

\paragraph{Abstract Type Members.} In CurrencyUnits.scala on line 6
there is an abstract type that is not supported. Note that we can not
simply replace it with a (supported) type parameter since the
CurrencyUnit class uses one of its implementing classes:
Satoshis. Since the Satoshis class overrides \texttt{A} with \texttt{Int64} anyway, we
just remove the abstract type declaration and replace \texttt{A} by \texttt{Int64}
everywhere.

\paragraph{Non-Literal BigInt Constructor Argument.} In
CurrencyUnits.scala on line 18 the BigInt constructor is called with a
non-literal argument. As described before, the types in the Stainless
library are more restricted than their Scala library counterparts. In
particular, the Stainless BigInt constructor is restricted to literal
arguments. So we simply replace \texttt{toLong} by
\texttt{underlying.toBigInt}: instead of converting the underlying
\texttt{Int64} (which in turn has an underlying \texttt{BigInt}) to
\texttt{Long} and then back to \texttt{BigInt} we simply directly
return the \texttt{BigInt}. This is an equivalent transformation: the
only thing that might go wrong in the detour via \texttt{Long} is that
the underlying \texttt{BigInt} does not fit into a
\texttt{Long}. However, the only constructor of \texttt{Int64Impl}
ensures exactly that and all functions producing \texttt{Int64} do so
via this constructor.

\paragraph{Self-Reference in Type Parameter Bound.} In
NumberTypes.scala both on lines 3 and 19 is a class with a type
parameter and a type boundary that contains that type parameter
itself. Stainless does not currently support such self-referential
type boundaries. We opened an issue \cite{Stainless:issue519} on the
Stainless repository and the developers have targeted version 0.4 to
support self-referential type boundaries. Since our code only uses
Number with type parameter \texttt{T} instantiated to \texttt{Int64}, we just remove the
type parameter declaration and replace all its occurrences by \texttt{Int64}.

\paragraph{Missing Member \texttt{bigInteger} in BigInt.} In
NumberType on line 6 there is a reference to \texttt{bigInteger}. The
Scala \texttt{BigInt} class is essentially a wrapper around
\texttt{java.math.BigInteger}. \texttt{BigInt} has a member
\texttt{bigInteger} which is the underlying instance of the Java
class. The Java class has a method \texttt{longValueExact} which
returns a \texttt{long} only if the \texttt{BigInteger} fits into a
\texttt{long}, otherwise throws exception. Stainless does not support
Java classes and in particular its \texttt{BigInt} has no member
\texttt{bigInteger}. However, our code does not call \texttt{toLong}
anymore, so we just remove it.

\paragraph{Type Members.} In NumberType.scala there is a type member
on line 4. Our version of Stainless (0.1) does not support type
members. We just remove the declaration and replace all occurrences of
\texttt{A} with \texttt{BigInt}, since \texttt{A} is never overwritten in an implementing class.
Note that in the meantime Stainless has implemented support for type
members \cite{Stainless:pull470}. Since version 0.2 verification
should succeed without this change.

\paragraph{Missing Bitwise-And Method on BigInt.} Contrary to Scala
\texttt{BigInt}, the Stainless \texttt{BigInt} class does not support bitwise
operations, in particular not the \&-method used in NumberType.scala
on line 13. However, as described above, the \texttt{checkResult} function is
both broken and redundant, so we remove it and all calls to it.

@@

\paragraph{Missing Implicit Long to BigInt Conversion.} The Scala
\texttt{BigInt} class has implict conversions from \texttt{Long} which
NumberType.scala uses on lines 32 and 34. They are missing in the
Stainless \texttt{BigInt}. A \texttt{BigInt} constructor with a
\texttt{Long} argument is also missing. We thus replace the
\texttt{Long} literals by an explicit call to the \texttt{BigInt}
constructor with a literal string argument,
e.g. \texttt{BigInt("-9223...5808")}.

@@

The original bitcoin-s code we started from, the extracted code, and
the finally verified code are available in our GitHub repository \cite{verification:github}.

@@
